---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bsimms

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/mattiaghilardi/bsimms/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mattiaghilardi/bsimms/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/mattiaghilardi/bsimms/graph/badge.svg)](https://app.codecov.io/gh/mattiaghilardi/bsimms)
<!-- [![CRAN status](https://www.r-pkg.org/badges/version/bsimms)](https://CRAN.R-project.org/package=bsimms) -->
<!-- badges: end -->

## Overview

bsimms is an R package for fitting Bayesian stable isotope mixing models
(SIMMs) via Stan, estimating the proportional contribution of two or more
sources to a mixture from tracer data such as stable isotopes. It brings
together mixing-model advances developed across the literature and
implemented in the JAGS-based package MixSIAR --- hierarchical source fitting,
fractionation error, concentration dependence, cross-tracer covariance,
flexible error structures --- and extends them by letting source
proportions depend on an arbitrary lme4-style fixed- and random-effects
formula of mixture-level covariates.

## Installation

You can install the development version of bsimms from GitHub like so:

``` r
# install.packages("pak")
pak::pak("mattiaghilardi/bsimms")
```

Fitting models also requires a working Stan backend, either
[cmdstanr](https://mc-stan.org/cmdstanr/) (recommended) or
[rstan](https://mc-stan.org/rstan/), neither of which is installed
automatically:

``` r
# cmdstanr (not on CRAN)
install.packages("cmdstanr", repos = c("https://stan-dev.r-universe.dev", getOption("repos")))
cmdstanr::check_cmdstan_toolchain()
cmdstanr::install_cmdstan()

# or rstan (on CRAN)
Sys.setenv(DOWNLOAD_STATIC_LIBV8 = 1) # only necessary for Linux without the nodejs library / headers
install.packages("rstan", repos = "https://cloud.r-project.org/", dependencies = TRUE)
```

See the [CmdStanR installation
guide](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) or the
[RStan Getting Started
guide](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started) for
OS-specific instructions and troubleshooting.

## How to use bsimms

As an example, we fit a mixing model to simulated data for a 3-source,
2-isotope model with 10 mixture samples, no covariates (`~1`), raw
replicate source data, and summarised (mean/SD) TDF data.

```{r sim-fit, message = FALSE, results = "hide"}
library(bsimms)

sim <- simulate_bsimms_data(
  ~1,
  n_mixture_obs = 10,
  source_names = c("Beaver", "Deer", "Hare"),
  isotope_names = c("d13C", "d15N"),
  seed = 1
)

fit <- bsimm(
  sim$formula,
  mixture_data = sim$mixture_data,
  source_data = sim$source_data,
  tdf_data = sim$tdf_data,
  isotope_names = sim$isotope_names,
  source_means_sds = sim$source_means_sds,
  tdf_means_sds = sim$tdf_means_sds,
  conc_dep = sim$conc_dep,
  error_structure = sim$error_structure,
  source_col = sim$source_col,
  cores = 4,
  seed = 1
)
```

Printing the fit gives a quick overview of the model.

```{r print-fit}
fit
```

`summary()` reports the population-average source proportions and error
term(s), each with a posterior mean, SD, 95% credible interval, and MCMC
convergence diagnostics (`rhat`, bulk/tail effective sample size). With the
default `error_structure = "process_residual"`, the error term shown per
isotope is `resid_prop`, MixSIAR's multiplicative factor scaling the
source/TDF variance propagated into the mixture -- not a raw residual SD.

```{r summary}
summary(fit)
```

`rhat` close to 1 and large effective sample sizes indicate the chains
mixed well. Population-average contributions lean towards Deer and Hare
over Beaver, though with fairly wide credible intervals given the small
simulated sample size.

Since this model has no covariates, every mixture sample shares the same
population-average proportions, so we can visualise their posterior
distribution directly.
  
```{r plot-proportions, out.width = "70%"}
p_arr <- posterior_proportions(fit)
plot_proportions(p_arr[, 1, , drop = FALSE], type = "density")
```
