---
title: "Modelling covariate effects"
author: "Mattia Ghilardi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Modelling covariate effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "covariates-",
  dev = "png",
  dpi = 150,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
```

## Overview

[`vignette("data-error")`](data-error.html) works through
**bsimms**'s core, MixSIAR-parity options -- data shapes, concentration
dependence, and error structures -- all fit with an intercept-only
(`~1`) formula. This vignette covers **bsimms**'s main extension
beyond MixSIAR, simmr, and cosimmr: an arbitrary `lme4`-style formula
of mixture-level covariates, spanning fixed effects (continuous,
factor, interaction) and crossed/nested random effects.

[`vignette("diagnostics")`](diagnostics.html) covers interpreting
Stan's sampler diagnostics.

## Fixed and random effects

Source proportions can depend on an `lme4`-style `formula` of
mixture-level covariates, added as deviations from the `p_global`
baseline in ILR space. Here we simulate a factor fixed effect (`Sex`),
a continuous fixed effect (`Age`), their interaction, and a random
intercept (`Region`):

```{r sim-covariates, message = FALSE, results = "hide"}
library(bsimms)

sim_cov <- simulate_bsimms_data(
  ~ Sex * Age + (1 | Region),
  n_mixture_obs = 60,
  source_names = c("Beaver", "Deer", "Hare"),
  isotope_names = c("d13C", "d15N"),
  n_levels = list(Sex = 2),
  n_groups = list(Region = 3),
  seed = 1
)

fit_full <- bsimm(
  sim_cov$formula,
  mixture_data = sim_cov$mixture_data,
  source_data = sim_cov$source_data,
  tdf_data = sim_cov$tdf_data,
  isotope_names = sim_cov$isotope_names,
  source_means_sds = sim_cov$source_means_sds,
  tdf_means_sds = sim_cov$tdf_means_sds,
  conc_dep = sim_cov$conc_dep,
  error_structure = sim_cov$error_structure,
  source_col = sim_cov$source_col,
  cores = 4,
  seed = 1
)
```

`summary()` now also reports fixed-effect coefficients (ILR-scale
deviations from `p_global`) and group-level standard deviations:

```{r summary-covariates}
summary(fit_full)
```

## Nested and crossed random effects

Nested random effects (e.g. sampling sites within regions) use
`lme4`'s `/` shorthand: `(1 | Region/Site)` expands to `(1 | Region) +
(1 | Site:Region)`, giving one random intercept per `Region` and a
further one per `Region`-`Site` combination. Crossed random effects,
where the grouping factors have no parent/child relationship (e.g.
`(1 | Sex:Region)`), use `:` the same way, without the implied
nesting. `chains`/`iter_warmup`/`iter_sampling` are reduced below
purely to keep this demonstration fast.

```{r sim-nested, message = FALSE, results = "hide"}
sim_nested <- simulate_bsimms_data(
  ~ (1 | Region / Site),
  n_mixture_obs = 60,
  source_names = c("Beaver", "Deer", "Hare"),
  isotope_names = c("d13C", "d15N"),
  n_groups = list(Region = 3, Site = 6),
  seed = 1
)

fit_nested <- bsimm(
  sim_nested$formula,
  mixture_data = sim_nested$mixture_data,
  source_data = sim_nested$source_data,
  tdf_data = sim_nested$tdf_data,
  isotope_names = sim_nested$isotope_names,
  source_means_sds = sim_nested$source_means_sds,
  tdf_means_sds = sim_nested$tdf_means_sds,
  conc_dep = sim_nested$conc_dep,
  error_structure = sim_nested$error_structure,
  source_col = sim_nested$source_col,
  chains = 2,
  iter_warmup = 500,
  iter_sampling = 500,
  seed = 1
)
```

`summary()` now reports a separate group-level SD for `Region` and
for `Site:Region` -- each region's own baseline deviation, and each
site's further deviation within its region. `n_groups$Site`'s total
(`6`) is split across `Region`'s levels the same way `n_mixture_obs`
is split across a factor's levels elsewhere, here 2 sites per region:

```{r summary-nested}
summary(fit_nested)
```

## Model comparison

To check whether these covariates are worth including, fit the same
data with the null (intercept-only) model and compare with
`loo_compare()`:

```{r loo-compare, message = FALSE, results = "hide"}
fit_null_cov <- bsimm(
  ~1,
  mixture_data = sim_cov$mixture_data,
  source_data = sim_cov$source_data,
  tdf_data = sim_cov$tdf_data,
  isotope_names = sim_cov$isotope_names,
  source_means_sds = sim_cov$source_means_sds,
  tdf_means_sds = sim_cov$tdf_means_sds,
  conc_dep = sim_cov$conc_dep,
  error_structure = sim_cov$error_structure,
  source_col = sim_cov$source_col,
  cores = 4,
  seed = 1
)
```

```{r loo-compare-result}
loo::loo_compare(fit_full, fit_null_cov)
```

The best-fitting model is listed first (`elpd_diff = 0`); a positive,
several-times-larger-than-its-SE `elpd_diff` for the runner-up would
indicate a real difference in predictive fit.

## Bayesian R^2^

With covariates that create real across-sample variation,
`rstantools::bayes_R2()` is now informative (unlike the
intercept-only case in
[`vignette("getting-started")`](getting-started.html), where it is
always ~0):

```{r bayes-r2-covariates}
rstantools::bayes_R2(fit_full)
```

R^2^ is reported per isotope, and relates directly to that isotope's
error term in `summary(fit_full)` above: `resid_prop` scales how much
source/TDF process variance is propagated into the mixture as
residual noise, so a smaller `resid_prop` -- more of that isotope's
variance explained by the covariates rather than left as noise --
corresponds to a higher R^2^, and vice versa, as seen here for `d15N`
(smaller `resid_prop`, higher R^2^) versus `d13C` (larger `resid_prop`,
lower R^2^).

## Conditional effects

`conditional_effects()` shows how source proportions vary across a
fixed-effect covariate, holding every other covariate at a reference
value and group-level terms at their population-average level. With
no `effects` argument, it computes one plot per term of the
fixed-effect formula -- main effects and two-way interactions alike:

```{r conditional-effects}
#| fig.alt:
#|   - >-
#|       Forest plot of posterior source proportions for Beaver, Deer,
#|       and Hare across the two levels of Sex, each with 50% and 95%
#|       credible intervals.
#|   - >-
#|       Ribbon plot of posterior source proportions for Beaver, Deer,
#|       and Hare across the range of Age, each with 50% and 95%
#|       credible intervals.
#|   - >-
#|       Forest plot of posterior source proportions for Beaver, Deer,
#|       and Hare across the two levels of Sex, faceted by three
#|       representative values of Age, each with 50% and 95% credible
#|       intervals.
#| fig.asp: 0.8
ce <- conditional_effects(fit_full)
plot(ce)
```

A factor covariate (`Sex`) is shown as a forest-plot-style linerange
at its levels; a numeric covariate (`Age`) as a ribbon across its
observed range. The interaction (`Sex:Age`) facets `Sex`'s linerange
plot by three representative values of `Age` (its mean and +/-1 SD by
default, overridable via `int_conditions`).

An `effects` entry names which covariate is swept (the first) and
which becomes the moderator (the second), so naming the same
interaction the other way round -- `"Age:Sex"` instead of `"Sex:Age"`
-- instead sweeps `Age` as a ribbon, faceted by `Sex`'s two levels:

```{r conditional-effects-swap}
#| fig.alt: >-
#|   Ribbon plot of posterior source proportions for Beaver, Deer, and
#|   Hare across the range of Age, faceted by the two levels of Sex,
#|   each with 50% and 95% credible intervals.
#| fig.asp: 0.8
ce_age <- conditional_effects(fit_full, effects = "Age:Sex")
plot(ce_age)
```

The random intercept on `Region` is, by default, marginalised out of
these plots (`re_formula = NA`); pass a reduced formula (e.g. `~ (1 |
Region)`) to condition on it instead, here or in
`posterior_proportions()`/`fitted_proportions()`/`predict()`.

## Visualising the random effect

`conditional_effects()` only ever sweeps a fixed-effect covariate, so
it cannot show `Region`'s own random effect. To visualise it instead,
predict directly for each of `Region`'s three fitted levels with
`posterior_proportions()`/`fitted_proportions()` and `newdata`,
holding `Sex`/`Age` at a reference value and conditioning
on `Region` via `re_formula`:

```{r region-newdata}
region_data <- data.frame(
  Sex = levels(sim_cov$mixture_data$Sex)[1],
  Age = mean(sim_cov$mixture_data$Age),
  Region = levels(sim_cov$mixture_data$Region)
)
region_data
```

```{r region-proportions, message = FALSE, results = "hide"}
p_region <- posterior_proportions(
  fit_full,
  newdata = region_data,
  re_formula = ~ (1 | Region)
)
```

`plot_proportions()` labels the x axis generically ("Observation"),
since the `[n_draws, n_obs, K]` array returned by
`posterior_proportions()` carries no information about what varies
between rows. It returns a plain `ggplot` object, though, so
relabelling it with `Region`'s own values needs only a couple of
`ggplot2` calls appended:

```{r plot-region}
#| fig.alt: >-
#|   Forest plot of posterior source proportions for Beaver, Deer,
#|   and Hare across the three levels of Region, each with 50% and
#|   95% credible intervals.
#| fig.asp: 0.8
plot_proportions(p_region, type = "interval") +
  ggplot2::labs(x = "Region") +
  ggplot2::scale_x_discrete(labels = region_data$Region)
```

## Predicting for unobserved data

The same `newdata` mechanism, introduced in
[`vignette("getting-started")`](getting-started.html), also predicts
for entirely new mixture samples -- including, here, a specific
individual (`Sex = "B"`, `Age = 2`) from a genuinely new, previously
unobserved `Region`. Set `allow_new_levels = TRUE` to sample a fresh
group-level deviation for the new region instead of erroring; with as
few levels as here, `sample_new_levels = "gaussian"` (drawing from
the estimated group-level SD) is more appropriate than the default
`"uncertainty"` (resampling one of the existing levels' draws):

```{r new-region, message = FALSE, results = "hide"}
new_region_data <- data.frame(Sex = "B", Age = 2, Region = "D")
p_new_region <- posterior_proportions(
  fit_full,
  newdata = new_region_data,
  re_formula = ~ (1 | Region),
  allow_new_levels = TRUE,
  sample_new_levels = "gaussian"
)
```

```{r plot-new-region}
#| fig.alt: >-
#|   Density plot of posterior source proportions for Beaver, Deer,
#|   and Hare for a new, previously unobserved Region.
#| fig.asp: 0.8
plot_proportions(p_new_region, type = "density")
```

The same `newdata`/`re_formula`/`allow_new_levels`/`sample_new_levels`
arguments work identically for `fitted()`/`predict()`, returning this
new individual's predicted mixture isotope values instead of its
source proportions. Once that individual is actually measured, this
prediction is a benchmark to flag a surprising sample: compare its
observed isotope values against it, the same idea as
[`vignette("getting-started")`](getting-started.html)'s posterior
predictive checks, but for one specific new sample rather than the
whole fitted dataset:

```{r new-region-predict}
predict(
  fit_full,
  newdata = new_region_data,
  re_formula = ~ (1 | Region),
  allow_new_levels = TRUE,
  sample_new_levels = "gaussian"
)
```
