---
title: "Specifying source data and error structures"
author: "Mattia Ghilardi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Specifying source data and error structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "data-error-",
  dev = "png",
  dpi = 150,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
```

## Overview

[`vignette("getting-started")`](getting-started.html) fits and
interprets a first **bsimms** model, using `bsimm()`'s default settings
with an intercept-only (`~1`) formula. For an individual mixture
sample $i$ and isotope $j$ (superscript $m$ for mixture, $s$ for
source, $t$ for trophic discrimination factor (TDF) throughout), the
expected value is

$$
\mu^m_{i, j} = \sum_{k=1}^{K} p_{i, k}
  \left(\mu^s_{k, j} + \mu^t_{k, j}\right),
  \qquad \sum_{k=1}^{K} p_{i, k} = 1, \ \ p_{i, k} > 0, \tag{1}
$$

where $p_{i, k}$ is source $k$'s proportional contribution to sample
$i$, and $\mu^s_{k, j}$ and $\mu^t_{k, j}$ are source $k$'s mean
isotope value and mean TDF for isotope $j$. The observed value is
normally distributed around it:

$$
y^m_{i, j} \sim \mathrm{Normal}\left(\mu^m_{i, j},
  \ {\sigma^m_{i, j}}^2\right). \tag{2}
$$

This vignette works through the modelling options that vary these
equations, beyond those defaults:

- raw vs. summarised source/TDF data
- concentration dependence
- the three error structures
- cross-tracer covariance, automatically estimated where applicable
  rather than a separate option to set

See [`vignette("covariates")`](covariates.html) for letting source
proportions depend on mixture-level covariates,
[`vignette("priors")`](priors.html) for **bsimms**'s data-scaled
default priors and how to override them, and
[`vignette("diagnostics")`](diagnostics.html) for interpreting Stan's
sampler diagnostics.

## Source and TDF data: raw vs. summarised

`source_data` and `tdf_data` can each be supplied in one of two shapes,
controlled by `source_means_sds`/`tdf_means_sds`:

```{r data-shapes}
# raw: one row per replicate measurement, one column per isotope
data.frame(
  Source = rep(c("Beaver", "Deer"), each = 2),
  d13C = c(-25.1, -24.8, -18.2, -17.9),
  d15N = c(5.1, 4.9, 8.2, 7.8)
)

# summarised: one row per source, mean/SD columns per isotope
data.frame(
  Source = c("Beaver", "Deer"),
  d13C_mean = c(-25, -18), d13C_sd = c(1, 1),
  d15N_mean = c(5, 8), d15N_sd = c(1, 1)
)
```

With raw data (`source_means_sds = FALSE`, the default), each
replicate measurement $r$ of source $k$'s isotope $j$ is
hierarchically fitted as

$$
y^s_{k, j, r} \sim \mathrm{Normal}\left(\mu^s_{k, j},
  \ {\sigma^s_{k, j}}^2\right), \tag{3}
$$

with $\mu^s_{k, j}$ from Equation (1) and $\sigma^s_{k, j}$ from the
error-structure equations below themselves estimated, so their
uncertainty is fully propagated into the source-proportion posterior.
With 2+ isotopes this is really the diagonal, no-correlation view of
a multivariate model (see [Cross-tracer
covariance](#cross-tracer-covariance) below). With summarised data
(`TRUE`), both are instead fixed at the supplied
values. `tdf_data` works the same way via `tdf_means_sds` (default
`TRUE`, since TDF/fractionation values are usually taken from the
literature as fixed mean/SD pairs, as in MixSIAR) -- **bsimms**
additionally extends this hierarchical fitting to raw TDF data, which
MixSIAR does not support:

$$
y^t_{k, j, r} \sim \mathrm{Normal}\left(\mu^t_{k, j},
  \ {\sigma^t_{k, j}}^2\right),
$$

with $\mu^t_{k, j}$ and $\sigma^t_{k, j}$ likewise estimated when
raw, from Equation (1) and the error-structure equations below
respectively.

## Concentration dependence

Concentration dependence ([Phillips & Koch
2002](https://doi.org/10.1007/s004420100786)) weights source
proportions by each source's elemental concentration for a given
tracer, rather than treating every source as contributing equally
regardless of concentration. It requires
an extra `<isotope>_conc` column per isotope in `source_data` (values in
`(0, 1]`, summing to at most 1 per source across isotopes), and is
enabled with `conc_dep = TRUE`. Formally, $p_{i, k}$ in Equation (1) is
replaced by a concentration-reweighted $p_{i, k}^*$:

$$
p_{i, k}^* = \frac{p_{i, k} \, c_{k, j}}{\sum_{l=1}^{K} p_{i, l} \, c_{l, j}},
$$

where $c_{k, j}$ is source $k$'s elemental concentration for isotope
$j$ (the `<isotope>_conc` column).

```{r conc-dep, message = FALSE, results = "hide"}
library(bsimms)

sim_conc <- simulate_bsimms_data(
  ~1,
  n_mixture_obs = 10,
  source_names = c("Beaver", "Deer", "Hare"),
  isotope_names = c("d13C", "d15N"),
  conc_dep = TRUE,
  seed = 1
)

fit_conc <- bsimm(
  sim_conc$formula,
  mixture_data = sim_conc$mixture_data,
  source_data = sim_conc$source_data,
  tdf_data = sim_conc$tdf_data,
  isotope_names = sim_conc$isotope_names,
  source_means_sds = sim_conc$source_means_sds,
  tdf_means_sds = sim_conc$tdf_means_sds,
  conc_dep = sim_conc$conc_dep,
  error_structure = sim_conc$error_structure,
  source_col = sim_conc$source_col,
  chains = 2,
  iter_warmup = 500,
  iter_sampling = 500,
  seed = 1
)
```

```{r conc-dep-print}
head(sim_conc$source_data)
fit_conc
```

## Error structures

`error_structure` controls ${\sigma^m_{i, j}}^2$ in Equation (2):

- `"process_residual"` (default): source/TDF variance is propagated into
  the mixture and scaled by an estimated `resid_prop` ($\phi_j$) factor
  -- MixSIAR's combined `"Residual * Process"` error structure ([Stock &
  Semmens 2016](https://doi.org/10.1002/ecy.1517)):
  ${\sigma^m_{i, j}}^2 = \phi_j \sum_{k=1}^{K} p_{i, k}^2
  \left({\sigma^s_{k, j}}^2 + {\sigma^t_{k, j}}^2\right)$.
- `"process_only"`: only source/TDF variance is propagated, with no
  further scaling, as in the original MixSIR model ([Moore & Semmens
  2008](https://doi.org/10.1111/j.1461-0248.2008.01163.x)):
  ${\sigma^m_{i, j}}^2 = \sum_{k=1}^{K} p_{i, k}^2
  \left({\sigma^s_{k, j}}^2 + {\sigma^t_{k, j}}^2\right)$.
- `"residual_only"`: source/TDF variance is not propagated into the
  mixture at all; all unexplained variance instead sits directly in
  ${\sigma^m_j}^2$ -- unlike the other two, not indexed by $i$, since
  it is fit directly rather than derived from $p_{i, k}$ -- an
  isotope-specific `sigma` term, as in the original SIAR model
  ([Parnell et al. 2010](https://doi.org/10.1371/journal.pone.0009672)).

```{r error-structure, message = FALSE, results = "hide"}
sim_resid <- simulate_bsimms_data(
  ~1,
  n_mixture_obs = 10,
  source_names = c("Beaver", "Deer", "Hare"),
  isotope_names = c("d13C", "d15N"),
  error_structure = "residual_only",
  seed = 1
)

fit_resid <- bsimm(
  sim_resid$formula,
  mixture_data = sim_resid$mixture_data,
  source_data = sim_resid$source_data,
  tdf_data = sim_resid$tdf_data,
  isotope_names = sim_resid$isotope_names,
  source_means_sds = sim_resid$source_means_sds,
  tdf_means_sds = sim_resid$tdf_means_sds,
  conc_dep = sim_resid$conc_dep,
  error_structure = sim_resid$error_structure,
  source_col = sim_resid$source_col,
  chains = 2,
  iter_warmup = 500,
  iter_sampling = 500,
  seed = 1
)
```

```{r error-structure-summary}
summary(fit_resid)
```

The error term is now `sigma` rather than `resid_prop`.

## Cross-tracer covariance

With 2+ isotopes, cross-tracer covariance ([Hopkins & Ferguson
2012](https://doi.org/10.1371/journal.pone.0028478)) generalises
Equation (2) to vector/matrix form:

$$
\mathbf{y}^m_i \sim
  \mathrm{MultivariateNormal}\left(\boldsymbol{\mu}^m_i, \ \Sigma^m_i\right),
$$

where $\Sigma^m_i$ is the $J \times J$ mixture covariance matrix for
sample $i$; its diagonal is ${\sigma^m_{i, j}}^2$ from the
error-structure equations above. When source data are raw, Equation
(3) likewise generalises:

$$
\mathbf{y}^s_{k, r} \sim
  \mathrm{MultivariateNormal}\left(\boldsymbol{\mu}^s_k, \ \Sigma^s_k\right),
$$

where $\Sigma^s_k$ is source $k$'s $J \times J$ covariance matrix; its
diagonal is ${\sigma^s_{k, j}}^2$ and its off-diagonal entries are
$\sigma^s_{k, j_1} \sigma^s_{k, j_2} \, \rho^s_{k, j_1, j_2}$, with
$\rho^s_{k, j_1, j_2}$ source $k$'s estimated correlation between
isotopes $j_1$ and $j_2$ (`bsimms_prior()`'s `"source_cor"` class).

Two specific combinations also estimate $\Sigma^m_i$'s off-diagonal,
cross-tracer entries automatically:

- Raw source data under a process-based error structure
  (`"process_only"`/`"process_residual"`): source correlation
  ($\Sigma^s_k$ above) propagates into $\Sigma^m_i$'s off-diagonal
  the same way its diagonal does, $\Sigma^m_i[j_1, j_2] =
  \sum_{k=1}^{K} p_{i, k}^2 \, \Sigma^s_k[j_1, j_2]$ for $j_1 \neq
  j_2$ (also scaled by $\sqrt{\phi_{j_1} \phi_{j_2}}$ under
  `"process_residual"`).
- `error_structure = "residual_only"`: a shared residual correlation,
  $\Sigma^m[j_1, j_2] = \sigma^m_{j_1} \sigma^m_{j_2} \,
  \rho^{\text{resid}}_{j_1, j_2}$ for $j_1 \neq j_2$, with $\Sigma^m$
  the shared $J \times J$ mixture covariance matrix (no $i$, as for
  $\sigma^m_j$ above) and $\rho^{\text{resid}}_{j_1, j_2}$ the
  estimated correlation between isotopes $j_1$ and $j_2$
  (`bsimms_prior()`'s `"resid_cor"` class).

Neither ever includes a TDF contribution, even when TDF data are raw
(matching MixSIAR, which also never modelled TDF cross-tracer
correlation).

Since source data are raw and there are two isotopes in `fit_resid`
above, its per-source (`Lcorr_source`) and shared residual
(`Lcorr_resid`) correlation Cholesky factors were also estimated;
both are ordinary Stan parameters, retrievable like any other via
`bsimms_draws()`, but are not part of `summary()`'s output since they
are a data/error-structure side effect rather than a quantity of
direct scientific interest here:

```{r cor-draws}
posterior::variables(bsimms_draws(fit_resid, variable = "Lcorr_source"))
posterior::variables(bsimms_draws(fit_resid, variable = "Lcorr_resid"))
```

## References

Phillips DL, Koch PL (2002). "Incorporating concentration dependence
in stable isotope mixing models." *Oecologia*, 130(1), 114-125.
[doi:10.1007/s004420100786](https://doi.org/10.1007/s004420100786)

Stock BC, Semmens BX (2016). "Unifying error structures in commonly
used biotracer mixing models." *Ecology*, 97(10), 2952-2960.
[doi:10.1002/ecy.1517](https://doi.org/10.1002/ecy.1517)

Moore JW, Semmens BX (2008). "Incorporating uncertainty and prior
information into stable isotope mixing models." *Ecology Letters*,
11(5), 470-480.
[doi:10.1111/j.1461-0248.2008.01163.x](https://doi.org/10.1111/j.1461-0248.2008.01163.x)

Parnell AC, Inger R, Bearhop S, Jackson AL (2010). "Source
partitioning using stable isotopes: coping with too much variation."
*PLoS ONE*, 5(3), e9672.
[doi:10.1371/journal.pone.0009672](https://doi.org/10.1371/journal.pone.0009672)

Hopkins JB, Ferguson JM (2012). "Estimating the diets of animals
using stable isotopes and a comprehensive Bayesian mixing model."
*PLoS ONE*, 7(1), e28478.
[doi:10.1371/journal.pone.0028478](https://doi.org/10.1371/journal.pone.0028478)
