---
title: "Getting started with bsimms"
author: "Mattia Ghilardi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with bsimms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "getting-started-",
  dev = "png",
  dpi = 150,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
```

## Overview

Fitting a mixing model with bsimms always needs the same five ingredients:

- `formula`: a one-sided `lme4`-style formula of mixture-level covariates
  on source proportions (`~1` for no covariates).
- `mixture_data`: a data frame with one row per mixture sample, a column
  per isotope, and, optionally, covariate columns used in `formula`.
- `source_data`: a data frame of isotope values for each candidate
  source, either raw replicate measurements or a summarised mean/SD per
  source (`source_means_sds = TRUE`).
- `tdf_data`: a data frame of the trophic discrimination factor (TDF,
  a.k.a. fractionation or enrichment factor) for each source, again raw
  or summarised (`tdf_means_sds`, `TRUE` by default).
- `isotope_names`: the isotope column names shared by `mixture_data`,
  `source_data`, and `tdf_data` above.

This vignette fits a model with no covariates end to end and works through
the results. See `vignette("model-specification")` for raw vs. summarised
source/TDF data, error structures, concentration dependence, and covariates,
and `vignette("diagnostics-troubleshooting")` for interpreting Stan's sampler
diagnostics and informational messages.

## Simulating example data

`simulate_bsimms_data()` generates data for any bsimms model configuration
from known parameter values, which is useful for learning the package (and
for testing). Here we simulate data for a 3-source, 2-isotope model with
10 mixture samples, no covariates, raw replicate source data without
concentration dependence (`conc_dep = FALSE`, the default), and summarised
(mean/SD) TDF data.

```{r sim, message = FALSE}
library(bsimms)

sim <- simulate_bsimms_data(
  ~1,
  n_mixture_obs = 10,
  source_names = c("Beaver", "Deer", "Hare"),
  isotope_names = c("d13C", "d15N"),
  seed = 1
)
```

`sim` bundles everything needed to fit the model (`mixture_data`,
`source_data`, `tdf_data`, and the settings below), plus `sim$truth`, the
true parameter values used to generate the data.

## Fitting the model

`bsimm()` builds the Stan program for the requested model, compiles it,
and draws posterior samples via MCMC. By default it uses whichever of
`cmdstanr`/`rstan` is installed (`cmdstanr` if both are); set `backend`
to choose explicitly. `chains = 4` MCMC chains are run by default; we
set `cores = 4` below so all four sample in parallel instead of one
after another.

```{r fit, message = FALSE, results = "hide"}
fit <- bsimm(
  sim$formula,
  mixture_data = sim$mixture_data,
  source_data = sim$source_data,
  tdf_data = sim$tdf_data,
  isotope_names = sim$isotope_names,
  source_means_sds = sim$source_means_sds,
  tdf_means_sds = sim$tdf_means_sds,
  conc_dep = sim$conc_dep,
  error_structure = sim$error_structure,
  source_col = sim$source_col,
  cores = 4,
  seed = 1
)
```

## Inspecting the fit

Printing the fit gives a one-screen overview of the model: the formula,
source/isotope/mixture counts and names, error structure, source/TDF data
mode, and fitting backend.

```{r print-fit}
fit
```

`summary()` reports the population-average source proportions and error
term(s), each with a posterior mean, SD, 95% credible interval, and MCMC
convergence diagnostics (`rhat`, bulk/tail effective sample size). With the
default `error_structure = "process_residual"`, the error term shown per
isotope is `resid_prop`, MixSIAR's multiplicative factor scaling the
source/TDF variance propagated into the mixture -- not a raw residual SD.

```{r summary}
summary(fit)
```

Population-average contributions lean towards Deer and Hare over Beaver,
though with fairly wide credible intervals given the small simulated
sample size. `rhat` close to 1 and large effective sample sizes indicate
the chains mixed well, which we can also check visually with trace and
density plots of the underlying Stan parameters.

```{r plot-fit}
#| fig.alt: "Density and trace plots of p_global and resid_prop across four MCMC chains; densities are unimodal and traces overlap with no trend, indicating good mixing."
#| fig.asp: 1.2
plot(fit)
```

## Posterior predictive checks

Beyond `rhat`/effective sample size, it is worth checking that the fitted
model actually reproduces the observed mixture data. `pp_check()` overlays
the observed isotope values (dark line) against posterior predictive
draws (light lines), one isotope at a time, using `bayesplot`'s
`dens_overlay` by default; pass `type` to use any other `bayesplot`
`ppc_*` function instead (e.g. `type = "ecdf_overlay"`). With only 10
mixture samples in this example, the observed density is naturally
rougher than it would be with more data -- judge the check by its
overall location and spread against the predictive draws, not by
matching every wiggle.

```{r pp-check}
#| fig.alt:
#|   - "Posterior predictive check for d13C: the observed density falls within the spread of posterior predictive draws."
#|   - "Posterior predictive check for d15N: the observed density falls within the spread of posterior predictive draws."
#| fig.asp: 0.8
bayesplot::pp_check(fit, resp = "d13C")
bayesplot::pp_check(fit, resp = "d15N")
```

## Working with posterior source proportions

`posterior_proportions()` returns the raw posterior draws of source
proportions `p` as a `[n_draws, n_obs, K]` array; `fitted_proportions()`
wraps it with the same summary measures as `summary()`. Since this model
has no covariates, every mixture sample shares the same population-average
proportions (`p_global`), so both return identical values for every
observation. Both also accept `newdata` to predict source proportions
for new, unobserved samples -- most useful once covariates are involved
(see `vignette("model-specification")`), since without them every new
sample would get the same population-average proportions shown here.

```{r fitted-proportions}
fitted_proportions(fit)
```

`plot_proportions()` visualises these draws. With a single set of
proportions shared by every mixture sample, a density/histogram plot of
population-average source proportions is the clearest view (`type =
"interval"` instead suits several distinct observations, e.g. one per
covariate level -- see `vignette("model-specification")`).

```{r plot-proportions-density}
#| fig.alt: "Density plot of posterior source proportions for Beaver, Deer, and Hare; Beaver's distribution is shifted towards lower proportions and more widely spread than the overlapping Deer and Hare distributions."
#| fig.asp: 0.8
p_arr <- posterior_proportions(fit)
plot_proportions(p_arr[, 1, , drop = FALSE], type = "density")
```

## Other posterior quantities

Besides source proportions, `fitted()`/`predict()` return the model's
expected/predicted mixture isotope values, useful for checking how well
the model reproduces the observed mixture data. As with source
proportions, every mixture sample shares the same expected isotope values
here, so we only show the first. `fitted()`/`predict()` accept the same
`newdata` argument as `posterior_proportions()`/`fitted_proportions()`
above, to predict mixture isotope values for new, unobserved samples.

```{r other-quantities}
subset(fitted(fit), row == 1)
```

## Bayesian R^2^

`rstantools::bayes_R2()` reports a Bayesian R-squared per isotope, but is
not useful with no covariates: every mixture sample shares the same
fitted value, so there is no across-sample variation for it to explain,
and it is always ~0. See `vignette("model-specification")`, where it
becomes informative once covariates are added.

## Next steps

- `vignette("model-specification")`: raw vs. summarised source/TDF data,
  error structures, concentration dependence, and covariates.
- `vignette("diagnostics-troubleshooting")`: interpreting Stan's sampler
  diagnostics and informational messages.
