---
title: "Getting started with bsimms"
author: "Mattia Ghilardi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with bsimms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "getting-started-",
  dev = "png",
  dpi = 150,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
```

Stable isotope mixing models (SIMMs) estimate the proportional
contribution of a set of sources to a mixture from differences in
stable isotope ratios (e.g. $\delta^{13}$C, $\delta^{15}$N) between
sources and the mixture -- most commonly for reconstructing animal
diets from tissue isotope values, though the same problem arises
elsewhere too, such as tracing the sources of pollutants or sediments.
The classic implementations, [SIAR](https://cran.r-project.org/package=siar)
([Parnell et al. 2010](https://doi.org/10.1371/journal.pone.0009672)) and
MixSIR ([Moore & Semmens
2008](https://doi.org/10.1111/j.1461-0248.2008.01163.x)), were later
combined into [MixSIAR](https://cran.r-project.org/package=MixSIAR)
([Stock et al. 2018](https://doi.org/10.7717/peerj.5096)), which -- together
with SIAR's successor [simmr](https://cran.r-project.org/package=simmr)
([Govan et al. 2023](https://doi.org/10.48550/arXiv.2306.07817)) -- remains
among the most widely used R packages for fitting Bayesian SIMMs in JAGS
today. [cosimmr](https://cran.r-project.org/package=cosimmr) ([Govan et al.
2024](https://doi.org/10.48550/arXiv.2408.17230)) extends simmr with
covariates, fit with a fast variational Bayes approximation instead of
JAGS.

**bsimms** fits this same class of models in Stan instead, using full
MCMC sampling. It reproduces MixSIAR's core modelling options -- raw
or summarised source and trophic discrimination factor (TDF) data,
concentration dependence, flexible error structures, and cross-tracer
covariance -- and extends them by letting source proportions depend
on an arbitrary `lme4`-style formula of fixed *and* random effects,
rather than MixSIAR's fixed menu of up to two categorical factors
plus one continuous covariate (cosimmr also supports covariates,
though via a fixed-effects design matrix only). Source
proportions are modelled in isometric log-ratio (ILR) coordinates, so
an unconstrained linear predictor maps onto the source simplex the
same way it would for any other response, via a numerically stable
softmax transform. It also uses weakly informative, data-scaled
default priors, robust to the skew and outliers common in isotope
data, with a full override system.

## Overview

Fitting a mixing model with **bsimms** always needs the same five ingredients:

- `formula`: a one-sided `lme4`-style formula of mixture-level covariates
  on source proportions (`~1` for no covariates).
- `mixture_data`: a data frame with one row per mixture sample, a column
  per isotope, and, optionally, covariate columns used in `formula`.
- `source_data`: a data frame of isotope values for each candidate
  source, either raw replicate measurements or a summarised mean/SD per
  source (`source_means_sds = TRUE`).
- `tdf_data`: a data frame of the trophic discrimination factor (TDF,
  a.k.a. fractionation or enrichment factor) for each source, again raw
  or summarised (`tdf_means_sds`, `TRUE` by default).
- `isotope_names`: the isotope column names shared by `mixture_data`,
  `source_data`, and `tdf_data` above.

This vignette fits a model with no covariates end to end and works
through the results. See [`vignette("data-error")`](data-error.html)
for raw vs. summarised source/TDF data, concentration dependence, and
error structures, [`vignette("covariates")`](covariates.html)
for letting source proportions depend on mixture-level covariates,
[`vignette("priors")`](priors.html) for **bsimms**'s data-scaled
default priors and how to override them, and
[`vignette("diagnostics")`](diagnostics.html) for interpreting Stan's
sampler diagnostics and informational messages.

## Worked example

### Simulating example data

`simulate_bsimms_data()` generates data for any **bsimms** model configuration
from known parameter values, which is useful for learning the package (and
for testing). Here we simulate data for a 3-source, 2-isotope model with
10 mixture samples, no covariates, raw replicate source data without
concentration dependence (`conc_dep = FALSE`, the default), and summarised
(mean/SD) TDF data.

```{r sim, message = FALSE}
library(bsimms)

sim <- simulate_bsimms_data(
  ~1,
  n_mixture_obs = 10,
  source_names = c("Beaver", "Deer", "Hare"),
  isotope_names = c("d13C", "d15N"),
  seed = 1
)
```

`sim` bundles everything needed to fit the model (`mixture_data`,
`source_data`, `tdf_data`, and the settings below), plus `sim$truth`, the
true parameter values used to generate the data.

### Fitting the model

`bsimm()` builds the Stan program for the requested model, compiles it,
and draws posterior samples via MCMC. By default it uses whichever of
`cmdstanr`/`rstan` is installed (`cmdstanr` if both are); set `backend`
to choose explicitly. `chains = 4` MCMC chains are run by default; we
set `cores = 4` below so all four sample in parallel instead of one
after another.

```{r fit, message = FALSE, results = "hide"}
fit <- bsimm(
  sim$formula,
  mixture_data = sim$mixture_data,
  source_data = sim$source_data,
  tdf_data = sim$tdf_data,
  isotope_names = sim$isotope_names,
  source_means_sds = sim$source_means_sds,
  tdf_means_sds = sim$tdf_means_sds,
  conc_dep = sim$conc_dep,
  error_structure = sim$error_structure,
  source_col = sim$source_col,
  cores = 4,
  seed = 1
)
```

### Inspecting the fit

Printing the fit gives a one-screen overview of the model: the formula,
source/isotope/mixture counts and names, error structure, source/TDF data
mode, and fitting backend.

```{r print-fit}
fit
```

`summary()` reports the population-average source proportions and error
term(s), each with a posterior mean, SD, 95% credible interval, and MCMC
convergence diagnostics (`rhat`, bulk/tail effective sample size). With the
default `error_structure = "process_residual"`, the error term shown per
isotope is `resid_prop`, MixSIAR's multiplicative factor scaling the
source/TDF variance propagated into the mixture -- not a raw residual SD.

```{r summary}
summary(fit)
```

Population-average contributions lean towards Deer and Hare over Beaver,
though with fairly wide credible intervals given the small simulated
sample size. `rhat` close to 1 and large effective sample sizes indicate
the chains mixed well, which we can also check visually with trace and
density plots of the underlying Stan parameters.

```{r plot-fit}
#| fig.alt: >-
#|   Density and trace plots of p_global and resid_prop across four
#|   MCMC chains; densities are unimodal and traces overlap with no
#|   trend, indicating good mixing.
#| fig.asp: 1.2
plot(fit)
```

### Posterior predictive checks

Beyond `rhat`/effective sample size, it is worth checking that the fitted
model actually reproduces the observed mixture data. `pp_check()` overlays
the observed isotope values (dark line) against posterior predictive
draws (light lines), one isotope at a time, using `bayesplot`'s
`dens_overlay` by default; pass `type` to use any other `bayesplot`
`ppc_*` function instead (e.g. `type = "ecdf_overlay"`). With only 10
mixture samples in this example, the observed density is naturally
rougher than it would be with more data -- judge the check by its
overall location and spread against the predictive draws, not by
matching every wiggle.

```{r pp-check}
#| fig.alt:
#|   - >-
#|       Posterior predictive check for d13C: the observed density
#|       falls within the spread of posterior predictive draws.
#|   - >-
#|       Posterior predictive check for d15N: the observed density
#|       falls within the spread of posterior predictive draws.
#| fig.asp: 0.8
bayesplot::pp_check(fit, resp = "d13C")
bayesplot::pp_check(fit, resp = "d15N")
```

### Working with posterior source proportions

`posterior_proportions()` returns the raw posterior draws of source
proportions `p` as a `[n_draws, n_obs, K]` array; `fitted_proportions()`
wraps it with the same summary measures as `summary()`. Since this model
has no covariates, every mixture sample shares the same population-average
proportions (`p_global`), so both return identical values for every
observation. Both also accept `newdata` to predict source proportions
for new, unobserved samples -- most useful once covariates are
involved (see [`vignette("covariates")`](covariates.html)), since
without them every new sample would get the same population-average
proportions shown here.

```{r fitted-proportions}
fitted_proportions(fit)
```

`plot_proportions()` visualises these draws. With a single set of
proportions shared by every mixture sample, a density/histogram plot of
population-average source proportions is the clearest view (`type =
"interval"` instead suits several distinct observations, e.g. one per
covariate level -- see [`vignette("covariates")`](covariates.html)).

```{r plot-proportions-density}
#| fig.alt: >-
#|   Density plot of posterior source proportions for Beaver, Deer,
#|   and Hare; Beaver's distribution is shifted towards lower
#|   proportions and more widely spread than the overlapping Deer and
#|   Hare distributions.
#| fig.asp: 0.8
p_arr <- posterior_proportions(fit)
plot_proportions(p_arr[, 1, , drop = FALSE], type = "density")
```

### Other posterior quantities

Besides source proportions, `fitted()`/`predict()` return the model's
expected/predicted mixture isotope values, useful for checking how well
the model reproduces the observed mixture data. As with source
proportions, every mixture sample shares the same expected isotope values
here, so we only show the first. `fitted()`/`predict()` accept the same
`newdata` argument as `posterior_proportions()`/`fitted_proportions()`
above, to predict mixture isotope values for new, unobserved samples.

```{r other-quantities}
subset(fitted(fit), row == 1)
```

### Bayesian R^2^

`rstantools::bayes_R2()` reports a Bayesian R-squared per isotope, but is
not useful with no covariates: every mixture sample shares the same
fitted value, so there is no across-sample variation for it to explain,
and it is always ~0. See [`vignette("covariates")`](covariates.html),
where it becomes informative once covariates are added.

## References

Parnell AC, Inger R, Bearhop S, Jackson AL (2010). "Source
partitioning using stable isotopes: coping with too much variation."
*PLoS ONE*, 5(3), e9672.
[doi:10.1371/journal.pone.0009672](https://doi.org/10.1371/journal.pone.0009672)

Moore JW, Semmens BX (2008). "Incorporating uncertainty and prior
information into stable isotope mixing models." *Ecology Letters*,
11(5), 470-480.
[doi:10.1111/j.1461-0248.2008.01163.x](https://doi.org/10.1111/j.1461-0248.2008.01163.x)

Stock BC, Jackson AL, Ward EJ, Parnell AC, Phillips DL, Semmens BX
(2018). "Analyzing mixing systems using a new generation of Bayesian
tracer mixing models." *PeerJ*, 6, e5096.
[doi:10.7717/peerj.5096](https://doi.org/10.7717/peerj.5096)

Govan E, Jackson AL, Inger R, Bearhop S, Parnell AC (2023). "simmr: A
package for fitting Stable Isotope Mixing Models in R." *arXiv
preprint*.
[doi:10.48550/arXiv.2306.07817](https://doi.org/10.48550/arXiv.2306.07817)

Govan E, Jackson AL, Bearhop S, Inger R, Stock BC, Semmens BX, Ward
EJ, Parnell AC (2024). "cosimmr: an R package for fast fitting of
Stable Isotope Mixing Models with covariates." *arXiv preprint*.
[doi:10.48550/arXiv.2408.17230](https://doi.org/10.48550/arXiv.2408.17230)
