---
title: "Setting custom priors"
author: "Mattia Ghilardi"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Setting custom priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "priors-",
  dev = "png",
  dpi = 150,
  fig.width = 6,
  out.width = "70%",
  fig.align = "center"
)
```

## Overview

[`vignette("data-error")`](data-error.html) and
[`vignette("covariates")`](covariates.html) work through **bsimms**'s
modelling options. This vignette covers its priors. Every parameter a
`bsimm()` model has -- source proportions, fixed- and random-effect
coefficients, error terms, and, when source/TDF data are raw, their
means/SDs -- gets a default prior automatically. Each can be
previewed with `bsimms_get_prior()` before fitting, and overridden,
in full or in part, with `bsimms_prior()`.

Some of these defaults are scaled to the data instead of fixed: the
residual error term under `error_structure = "residual_only"`
(`sigma`), and, when source/TDF data are raw, each source's own mean
and SD (`source_mean`/`source_sd`, `tdf_mean`/`tdf_sd`). These use the
sample median and MAD (median absolute deviation) of the relevant
data, rather than a single fixed default shared across every model
and dataset. Median and MAD are robust to the skew and outliers
common in isotope data.

## Previewing default priors

`bsimms_get_prior()` previews the default priors **bsimms** would use
for a given model configuration, without fitting it:

```{r sim, message = FALSE}
library(bsimms)

sim <- simulate_bsimms_data(
  ~1,
  n_mixture_obs = 10,
  source_names = c("Beaver", "Deer", "Hare"),
  isotope_names = c("d13C", "d15N"),
  seed = 1
)

bsimms_get_prior(
  ~1,
  mixture_data = sim$mixture_data,
  source_data = sim$source_data,
  tdf_data = sim$tdf_data,
  isotope_names = sim$isotope_names,
  error_structure = sim$error_structure
)
```

## Overriding priors

Any row can be overridden with `bsimms_prior()`, passed to `bsimm()`'s
`prior` argument. A more specific row (naming a `coef`/`resp`/`group`)
takes precedence over a general `class` default, and a general
override cascades onto every more specific default row it would
otherwise be shadowed by, rather than being silently unused.
`bsimms_get_prior()` above pre-fills one `p_global` row per source; a
general override with no `group` applies to all three at once:

```{r prior-override-general, message = FALSE, results = "hide"}
fit_general <- bsimm(
  sim$formula,
  mixture_data = sim$mixture_data,
  source_data = sim$source_data,
  tdf_data = sim$tdf_data,
  isotope_names = sim$isotope_names,
  source_means_sds = sim$source_means_sds,
  tdf_means_sds = sim$tdf_means_sds,
  conc_dep = sim$conc_dep,
  error_structure = sim$error_structure,
  source_col = sim$source_col,
  prior = bsimms_prior("2", class = "p_global"),
  chains = 2,
  iter_warmup = 500,
  iter_sampling = 500,
  seed = 1
)
```

```{r prior-override-general-check}
fit_general$prior[fit_general$prior$class == "p_global", ]
```

A `group`-specific override instead only replaces that one source's
row, taking precedence over the general one.
For example, to additionally give Beaver's share a
stronger-than-default concentration:

```{r prior-override, message = FALSE, results = "hide"}
fit <- bsimm(
  sim$formula,
  mixture_data = sim$mixture_data,
  source_data = sim$source_data,
  tdf_data = sim$tdf_data,
  isotope_names = sim$isotope_names,
  source_means_sds = sim$source_means_sds,
  tdf_means_sds = sim$tdf_means_sds,
  conc_dep = sim$conc_dep,
  error_structure = sim$error_structure,
  source_col = sim$source_col,
  prior = bsimms_prior("3", class = "p_global", group = "Beaver"),
  chains = 2,
  iter_warmup = 500,
  iter_sampling = 500,
  seed = 1
)
```

```{r prior-override-check}
fit$prior[fit$prior$class == "p_global" & fit$prior$group == "Beaver", ]
```

Several overrides can be combined with `c()`, e.g. to also raise
Deer's concentration at the same time:

```{r prior-override-multi}
c(
  bsimms_prior("3", class = "p_global", group = "Beaver"),
  bsimms_prior("2", class = "p_global", group = "Deer")
)
```

Other parameter classes work the same way. `bsimms_get_prior()`'s
output above shows `resid_prop`'s default is `uniform(0, 20)` for
both isotopes; `resid_prop` is restricted to non-negative values, so a
custom prior needs a matching, positive-only distribution -- e.g. a
`gamma(2, 2)`, centred near its expected value of 1, for `d13C`:

```{r prior-override-resid-prop, message = FALSE, results = "hide"}
fit_resid_prop <- bsimm(
  sim$formula,
  mixture_data = sim$mixture_data,
  source_data = sim$source_data,
  tdf_data = sim$tdf_data,
  isotope_names = sim$isotope_names,
  source_means_sds = sim$source_means_sds,
  tdf_means_sds = sim$tdf_means_sds,
  conc_dep = sim$conc_dep,
  error_structure = sim$error_structure,
  source_col = sim$source_col,
  prior = bsimms_prior("gamma(2, 2)", class = "resid_prop", resp = "d13C"),
  chains = 2,
  iter_warmup = 500,
  iter_sampling = 500,
  seed = 1
)
```

```{r prior-override-resid-prop-check}
fit_resid_prop$prior[fit_resid_prop$prior$class == "resid_prop", ]
```

### Informative priors from independent source-contribution data

When independent information on each source's contribution already
exists -- from a previous study, a different data type, or expert
knowledge -- it can be turned into an informative Dirichlet prior for
every source's `p_global` at once. Rescale the raw counts or
percentages so they sum to $K$, the number of sources ([Stock et al.
2018](https://doi.org/10.7717/peerj.5096), eq. 4): this keeps the
prior's total weight the same as the default, "generalist"
$\mathrm{Dirichlet}(1, \ldots, 1)$ prior, while preserving the
independent data's relative proportions among sources.

A common example is diet composition estimated from stomach or
faecal content analysis, but the same approach applies to any
source-contribution estimate from outside the isotope data itself --
e.g. known upstream contributions in a sediment- or
pollutant-tracing application:

```{r informative-prior-alpha}
# hypothetical stomach-content counts for each source
diet_counts <- c(Beaver = 30, Deer = 18, Hare = 12)

# rescale so the concentrations sum to K (Stock et al. 2018, eq. 4)
alpha <- length(diet_counts) * diet_counts / sum(diet_counts)
alpha
```

```{r informative-prior-fit, message = FALSE, results = "hide"}
informative_prior <- c(
  bsimms_prior("1.5", class = "p_global", group = "Beaver"),
  bsimms_prior("0.9", class = "p_global", group = "Deer"),
  bsimms_prior("0.6", class = "p_global", group = "Hare")
)

fit_informative <- bsimm(
  sim$formula,
  mixture_data = sim$mixture_data,
  source_data = sim$source_data,
  tdf_data = sim$tdf_data,
  isotope_names = sim$isotope_names,
  source_means_sds = sim$source_means_sds,
  tdf_means_sds = sim$tdf_means_sds,
  conc_dep = sim$conc_dep,
  error_structure = sim$error_structure,
  source_col = sim$source_col,
  prior = informative_prior,
  chains = 2,
  iter_warmup = 500,
  iter_sampling = 500,
  seed = 1
)
```

```{r informative-prior-check}
fit_informative$prior[fit_informative$prior$class == "p_global", ]
```

## References

Stock BC, Jackson AL, Ward EJ, Parnell AC, Phillips DL, Semmens BX
(2018). "Analyzing mixing systems using a new generation of Bayesian
tracer mixing models." *PeerJ*, 6, e5096.
[doi:10.7717/peerj.5096](https://doi.org/10.7717/peerj.5096)
